---
title: "8: Testing"
subtitle: "R for faculty, Fall 2024"
author: "Jonas Moss"
format:
  revealjs:
    error: True
    theme: [default, custom.scss]
    width: 1600
    height: 900
    embed-resources: True
    mainfont: "Verdana"
    #min-scale: 0.2
    #max-scale: 2
    echo: True
    auto-stretch: False
---

# Testing

## Install the `datarium` package

If you haven't already, install the package.
```{r}
#| eval: False
install.packages("datarium")
```


## Paired *t*-test

::: {.panel-tabset}

### Data

```{r}
mice2 = tibble::as_tibble(datarium::mice2)
mice2
```

### The paired *t*-test

```{r}
mice2_test = t.test(mice2$before, mice2$after, paired = TRUE)
mice2_test
```

### Exercise
Use the `str` function to inspect `mice2_test`. Extract the *p*-value. Why would you do this?

### Solution
```{r}
str(mice2_test)
```

```{r}
mice2_test$p.value
```

:::

## Two-sample *t*-test

::: {.panel-tabset}

### Data
```{r}
weight = datarium::genderweight
head(weight)
```


### Test
```{r}
females = weight$weight[weight$group == "F"]
males = weight$weight[weight$group == "M"]
ts_test = t.test(males, females)
```

### Exercise
Use `str` to inspect; alternatively, use the console. Find the *t*-value of the test. (Often reported in, e.g., psychology.)

### Solution
```{r}
ts_test$statistic
```


:::


## Chi square test

::: {.panel-tabset}

### Data
```{r}
properties = datarium::properties
head(properties)
```

### Table

Are `property_type` and `buyer_type` independent?

```{r}
tab <- table(properties$property_type, properties$buyer_type)
tab
```

### `chisq_test``
```{r}
chisq_test <- chisq.test(tab)
chisq_test
```

### Exercise
Find out how to access the *p*-value and `X-squared` of the preceeding test. Store them in the variables `pval` and `x_squared`.

### Solution
```{r}
# str(chisq_test)
pval <- chisq_test$p.value
x_squared <- chisq_test$statistic
```
:::


## Correlation test

```{r}
cor.test(mice2$after, mice2$before)
```

::: {.panel-tabset}
### Exercise
Using the `datarium::anxiety` data set, test if the correlation between `t1` and `t2` is significantly different from `0`. Store the resulting *p*-value inside the variable `pval`. 

### Solution
```{r}
val <- cor.test(datarium::anxiety$t1, datarium::anxiety$t2)
pval <- val$p.value
```
:::


## Kolmogorov--Smirmov test

::: {.panel-tabset}

### Data
```{r}
jobsatisfaction <- datarium::jobsatisfaction
summary(jobsatisfaction)
```

### Histogram
Are the distributions different?
```{r}
male_scores <- jobsatisfaction$score[jobsatisfaction$gender == "male"]
female_scores <- jobsatisfaction$score[jobsatisfaction$gender == "female"]
hist(male_scores, col = rgb(0, 0, 1, 0.5), main = "Job satisfaction")
hist(female_scores, col = rgb(1, 0, 0, 0.5), add = TRUE)
```

### ECDF
Are the distributions different? (ECDF: Empirical cumulative distribution)
```{r}
male_scores <- jobsatisfaction$score[jobsatisfaction$gender == "male"]
female_scores <- jobsatisfaction$score[jobsatisfaction$gender == "female"]
plot(ecdf(male_scores), col = "blue", main = "Job satisfaction")
plot(ecdf(female_scores), col = "red", add = TRUE)
```

### KS test
```{r}
ks.test(male_scores, female_scores)
```
### Exercise
Run a *t*-test to figure out if male job satisfaction is significantly different from female job satisfaction.

### Solution 
```{r}
t.test(male_scores, female_scores)
```

### Exercise
Are the distributions of the scores significantly different for the two levels of education? Use a plot and the Kolomogorov--Smirnov test.

### Solution
```{r}
education <- jobsatisfaction$education_level
school_score <- jobsatisfaction$score[education == "school"]
college_score <- jobsatisfaction$score[education == "college"]
plot(ecdf(school_score), col = "blue", main = "Job satisfaction", xlim = c(4, 7))
plot(ecdf(college_score), col = "red", add = TRUE)
```
```{r}
ks.test(school_score, college_score)
```

:::

# Summary

1. `R` supports most basic tests.
2. More specialized tests, e.g., the Dickey--Fuller test for unit roots is in the `tseries` library.
3. Inspect the test objects to find the information you need.

. . . 

* Some external libraries are, of course, unreliable. 
* Others are not ideally constructed, so that you cannot access the information you need. But this is rate.

